Linear Algebra as a Unifying Theory\\of Wireless Workload Modeling

### 2008-02-02 paper.tex
More work on explaining flow behavior.

\subsection{Flow Behavior}

The behavior of a flow may be mapped into a vector space such that the two operations, namely vector addition and scalar multiplication, can be sensibly defined in a way that naturally corresponds with the aggregation of behaviors. This is best clarified with an intentionally simplistic example. Suppose that the behavior of a flow is represented by the number of packets, $p$, the total bytes transmitted, $b$, and the total duration, $d$. This triplet can be naturally embedded in Euclidean space: $\vec{x}=(p,b,d)\in\R^3$. If we have two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, then their ``sum'' behavior, $\vec{x}_1+\vec{x}_2$ has $p_1+p_2$ packets, $b_1+b_2$ bytes, $d_1+d_2$ duration. Their average, i.e. $\frac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.

@@

The {packets-bytes-duration} model representation of flow behavior is clearly quite simplistic. It captures nothing about variability in packet sizes, nor about the ``burstiness'' of packet transmissions. At the opposite extreme, it is possible to embed the exact start time of each flow, and the exact size and transmission times of its packets in a vector space. The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to aggregate behavior of flows. Instead, we take an intermediate approach, suggested by the main result of~\cite{Karpinski07:realism}: if the start time and end time of each flow is known, it suffices to know the empirical distribution of packet sizes and inter-packet intervals in order to reproduce network performance characteristics accurately. Accordingly, we represent a flow's behavior by these precise characteristics.

Explicitly, we encode a flow's behavior with four vectors: $\vec{z} \in \R^{d_z}$ is a histogram of packet sizes; $\vec{v} \in \R^{d_v}$ is a histogram of inter-packet intervals; $\vec{\alpha}, \vec{\epsilon} \in \R^{d_t}$ are vectors indicating when the flow begins and ends. To expand on this, $z_i$ (i.e. the $i$th element of $\vec{z}$) is the number of packets in the flow such with $i$ bytes and $d_z$ is the largest possible packet size, which in practice is the maximum transfer unit of the wireless medium---typically 1500 bytes. The histogram of inter-packet intervals is similar, except that domain of possible intervals is a continuum, so we much map the range of possible intervals onto the discrete index values $\left\{1,\dots,d_v\right\}$. In our experiments, we use the following logarithmic quantization function:
\begin{align}
q(v) &= \left\lfloor(d_v-1)\frac{\ln v - \ln v_{\min}}{\ln v_{\max}-\ln v_{\min}}\right\rfloor+1.
\end{align}
with $v_{\min}=10^{-6}$, which is the minimal interval our trace data, and $v_{\max}=600$, which is the total duration of each simulated situation, and hence an upper bound on the inter-packet intervals. We use $d_v=1500$ to match the dimensionally of the packet size representation.

The flow start and end time indicator vectors, $\vec{\alpha}$ and $\vec{\epsilon}$, also require some explanation. Start and end times are simple scalar values. Why not represent them as such? The answer lies in the requirement that behaviors must add meaningfully. If we represent the start time as a scalar, then adding the behavior of two flows produces a (fairly) meaningless value. It's certainly not a start time itself---it may even exceed the start time of any actual flow. Instead, we again quantize the continuum of time: we map the range of possible start times onto the indices $\left\{1,\dots,d_t\right\}$, where $d_t$ is some chosen dimensionality. If the start time is mapped to index, $i$, then $\vec{\alpha}=e_i$, the vector with 1 at index $i$ and zeros elsewhere. The benefit of this representation is that we can now appropriately represent aggregate behavior through vector addition: the sum of start vectors over any collection of flows represents is a histogram of start times for those flows.

Finally, we collect all the pieces of flow behavior into matrix representations of behavior across an entire traffic trace. If there are $f$ flows in the trace, we define the size matrix $Z\in\R^{f \times d_z}$ to be the matrix whose row vectors are the size vectors $\vec{z}_i$ for $i=1,\dots,f$. In precisely the same manner, we define the interval matrix, $V\in\R^{f \times d_v}$, and the start time matrix, $A\in\R^{f \times d_t}$, and the end time matrix, $E\in\R^{f \times d_t}$.

\subsection{Flow Sources \& Destinations}

>>>>

The traffic workload pattern in a network is simply a collection of sequences of packets between hosts in the network sent at certain times. It is standard in network analysis to aggregate sequences of packets sharing the same \caps{IP} protocol, source and destination nodes, and source and destination \caps{TCP/UDP} port numbers. Such a sequence of packets is called a \textit{flow}. The traffic pattern is characterized completely by the detailed behavior of it's flows. The behavior of a is defined by six properties: its source and destination nodes, its start and end times, and the sizes of and intervals between its constituent packets. We propose to represent the behavior of each flow in terms of six vectors---one for each aspect of the flow's behavior. The aggregate behavior of all the flows in a workload pattern may then be represented by six matrices, each row of which gives the behavior vectors of one of the flows.

When mapping flow behaviors into a vector space, it is important that the mapping of behaviors into the vector spaces in question respect the vector space operations: addition of vectors and scalar multiplication. This requirement is best clarified with an intentionally simplistic example. Suppose that the behavior of a flow is represented by the number of packets, $p$, the total bytes transmitted, $b$, and the total duration, $d$. This triplet can be naturally embedded into Euclidean space: $\vec{x}=(p,b,d)\in\R^3$. If we have two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, then their ``sum'' behavior, $\vec{x}_1+\vec{x}_2$ has $p_1+p_2$ packets, $b_1+b_2$ bytes, $d_1+d_2$ duration. Their average, i.e. $\frac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.

%The behavior of a flow may be mapped into a vector space such that the two operations, namely vector addition and scalar multiplication, naturally correspond with the aggregation of behaviors. This requirement is best clarified with an intentionally simplistic example. Suppose that the behavior of a flow is represented by the number of packets, $p$, the total bytes transmitted, $b$, and the total duration, $d$. This triplet can be naturally embedded into Euclidean space: $\vec{x}=(p,b,d)\in\R^3$. If we have two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, then their ``sum'' behavior, $\vec{x}_1+\vec{x}_2$ has $p_1+p_2$ packets, $b_1+b_2$ bytes, $d_1+d_2$ duration. Their average, i.e. $\frac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.

@@

The {packets-bytes-duration} model representation is clearly quite simplistic, not to mention incomplete. It captures nothing about variability in packet sizes, nor about the ``burstiness'' of packet transmissions; it leaves start time unspecified; and it says nothing about the source or destination of the flow. At the opposite extreme, it is possible to embed the exact start time of each flow, and the exact size and transmission times of its packets in a vector space. The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to aggregate behavior of flows. We take an intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: if the end-points, start time and end time of each flow is known, it suffices to know the empirical distribution of packet sizes and inter-packet intervals in order to reproduce network performance characteristics accurately. Accordingly, we represent a flow's behavior by these precise characteristics. We encode a single flow's behavior with six component vectors:
\begin{itemize}
\item $\vec{z} \in \R^{d_z}$ is a histogram of packet sizes;
\item $\vec{v} \in \R^{d_v}$ is a histogram of inter-packet intervals;
\item $\vec{b} \in \R^{d_t}$ indicates when the flow begins;
\item $\vec{e} \in \R^{d_t}$ indicates when the flow ends;
\item $\vec{s}\in\R^n$ indicates the source node of the flow;
\item $\vec{d}\in\R^n$ indicates the destination node of the flow.
\end{itemize}
The behavior vectors for all flows are collected into six corresponding matrices, whose rows are the behavior vectors above:
\begin{itemize}
\item $Z\in\R^{f \times d_z}$ the size matrix;
\item $V\in\R^{f \times d_v}$ the interval matrix;
\item $B\in\R^{f \times d_t}$ the start time matrix;
\item $E\in\R^{f \times d_t}$ the end time matrix;
\item $S\in\R^{f \times n}$ the source matrix;
\item $D\in\R^{f \times n}$ the destination matrix.
\end{itemize}

%We collect all the components of flow behavior into matrix representations of behavior across an entire traffic trace. If there are $f$ flows in the trace, we define the size matrix $Z\in\R^{f \times d_z}$ to be the matrix whose rows are the size vectors $\vec{z}_i$ for $i=1,\dots,f$. In precisely the same manner, we define the interval matrix, $V\in\R^{f \times d_v}$, and the start time matrix, $A\in\R^{f \times d_t}$, and the end time matrix, $E\in\R^{f \times d_t}$.

\subsection{Packet Sizes \& Inter-Packet Intervals}

To expand on this, $z_i$ (i.e. the $i$th element of $\vec{z}$) is the number of packets in the flow such with $i$ bytes and $d_z$ is the largest possible packet size, which in practice is the maximum transfer unit of the wireless medium---typically 1500 bytes. The histogram of inter-packet intervals is similar, except that domain of possible intervals is a continuum, so we much map the range of possible intervals onto the discrete index values $\left\{1,\dots,d_v\right\}$. In our experiments, we use the following logarithmic quantization function:
\begin{align}
q(v) &= \left\lfloor(d_v-1)\frac{\ln v - \ln v_{\min}}{\ln v_{\max}-\ln v_{\min}}\right\rfloor+1.
\end{align}
with $v_{\min}=10^{-6}$, which is the minimal interval our trace data, and $v_{\max}=600$, which is the total duration of each simulated situation, and hence an upper bound on the inter-packet intervals. We use $d_v=1500$ to match the dimensionally of the packet size representation. The packet size and inter-packet intervals are collected into matrices, $Z\in\R^{f \times d_z}$ and $V\in\R^{f \times d_v}$, whose rows are the behavior vectors $\vec{z}_i$ and $\vec{v}_i$ indexed from $i=1,\dots,f$ where $f$ is the number of flows in the collection.

\subsection{Start \& End Times}

The flow start and end time indicator vectors, $\vec{\alpha}$ and $\vec{\epsilon}$, also require some explanation. Start and end times are simple scalar values. Why not represent them as such? The answer lies in the requirement that behaviors must add meaningfully. If we represent the start time as a scalar, then adding the behavior of two flows produces a meaningless value. It's certainly not a start time itself---it may even exceed the start time of any actual flow. Instead, we again quantize the continuum of time: we map the range of possible start times onto the indices $\left\{1,\dots,d_t\right\}$, where $d_t$ is some chosen dimensionality. If the start time is mapped to index, $i$, then $\vec{\alpha}=e_i$, the vector with 1 at index $i$ and zeros elsewhere. The benefit of this representation is that we can now appropriately represent aggregate behavior through vector addition: the sum of start vectors over any collection of flows is a histogram of start times for those flows.

\subsection{Source \& Destination Nodes}
